A Formal Treatment of Bidirectional Typing

### 2024-01-22 tex/ESOP24/S6-example.tex
[ tex ] Camera ready

\ifarxiv
For illustrative purposes, consider a computational calculus~\cite{Moggi1989} with additional constructs listed in \cref{tab:computational-calculus}.
The extended language has `only' 15 constructs, far fewer than a realistic programming language would have, but there are nearly 100 possible cases to consider in bidirectional type synthesis.

On the other hand, similar to a parser generator, a type-synthesiser generator only needs a specification $(\Sigma, \Omega)$ from the user to produce a corresponding synthesiser. 
Such a specification can be derived by extending $\Sigma_{\bto}$ with additional type constructs with $\mathsf{nat}$, $\mathsf{prod}$, $\mathsf{sum}$, and $\mathsf{T}$ such that
\[
  \arity({\mathsf{nat}}) = 0\qquad
  \arity({\mathsf{T}}) = 1\qquad
  \arity({\mathsf{prod}}) = \arity({\mathsf{sum}}) = 2.
\]
Types $\mathtt{nat}$, $A \times B$, $A + B$, and $T(A)$ are given as $\tyOp_{\mathsf{nat}}$, $\tyOp_{\mathsf{prod}}(A, B)$, $\tyOp_{\mathsf{sum}}(A, B)$, and $\tyOp_{\mathsf{T}}(A)$ respectively.
The signature $\Omega_{\Lambda}^{\Leftrightarrow}$ is then extended with operations listed in \cref{tab:computational-calculus}.
Mode-correctness can be derived by \cref{lem:decidability-mode-correctness} and its type synthesiser by \cref{cor:trichotomy} with the specification~$(\Sigma, \Omega)$ directly.
For a demonstration in \Agda, see \cref{sec:formalisation}.
\else
For illustrative purposes, consider a computational calculus~\cite{Moggi1989} with naturals, sums, products, and general recursion.
The extended language has `only' 15 constructs, including pairing, projections, injections, and so on, and this number of constructs is still far fewer than what a realistic programming language would have.
Even for this small calculus, there are already nearly 100 possible cases to consider in bidirectional type synthesis.

On the other hand, similar to a parser generator, only one specification is needed for a type-synthesiser generator from the user to produce a corresponding type synthesiser. 
Such a specification can be derived by extending $(\Sigma_{\bto}, \Omega_{\Lambda}^{\Leftrightarrow})$ accordingly for additional types and constructs with mode-correctness proved by applying \cref{lem:decidability-mode-correctness}, so its type synthesiser follows from \cref{cor:trichotomy} directly.
\fi

>>>>

\ifarxiv
For illustrative purposes, consider a computational calculus~\cite{Moggi1989} with additional constructs listed in \cref{tab:computational-calculus}.
\else
For illustrative purposes, consider a computational calculus~\cite{Moggi1989} with naturals, sums, products, and general recursion.
\fi
The extended language has `only' 15 constructs, including pairing, projections, injections, and so on, and this number of constructs is still far fewer than what a realistic programming language would have.
Even for this small calculus, there are already nearly 100 possible cases to consider in bidirectional type synthesis.

On the other hand, similar to a parser generator, only one specification is needed for a type-synthesiser generator from the user to produce a corresponding type synthesiser. 
\ifarxiv
Such a specification can be derived by extending $\Sigma_{\bto}$ with additional type constructs with $\mathsf{nat}$, $\mathsf{prod}$, $\mathsf{sum}$, and $\mathsf{T}$ such that
\[
  \arity({\mathsf{nat}}) = 0\qquad
  \arity({\mathsf{T}}) = 1\qquad
  \arity({\mathsf{prod}}) = \arity({\mathsf{sum}}) = 2.
\]
Types $\mathtt{nat}$, $A \times B$, $A + B$, and $T(A)$ are given as $\tyOp_{\mathsf{nat}}$, $\tyOp_{\mathsf{prod}}(A, B)$, $\tyOp_{\mathsf{sum}}(A, B)$, and $\tyOp_{\mathsf{T}}(A)$ respectively.
The signature $\Omega_{\Lambda}^{\Leftrightarrow}$ is then extended with operations listed in \cref{tab:computational-calculus}.
Mode-correctness can be proved by applying \cref{lem:decidability-mode-correctness}, so its type synthesiser follows from \cref{cor:trichotomy} with the extended specification.
For a demonstration in \Agda, see \cref{sec:formalisation}.
\else
Such a specification can be derived by extending $(\Sigma_{\bto}, \Omega_{\Lambda}^{\Leftrightarrow})$ accordingly for additional types and constructs with mode-correctness proved by applying \cref{lem:decidability-mode-correctness}, so its type synthesiser follows from \cref{cor:trichotomy} directly.
\fi

### 2024-01-22 tex/ESOP24/S8-future.tex
[ tex ] Camera ready

\paragraph{Dependent types}
In the context of dependent type theory, type synthesis entails a term equality check for type equality.
Achieving decidable bidirectional type synthesis depends on the decidability of term equality, which, in turn, relies on the ability to specify computation rules within a language specification.

Extending the logical framework \Dedukti~\cite{Assaf2016}, \varcitet{Felicissimo2023}{'s recent \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
They propose a notion of \LF signature with modes, a definition of mode-correctness, and an algorithm of bidirectional type synthesis.
The algorithm is decidable, sound and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and that the set of computational rules is type-preserving, confluent, and normalising.
The last assumption, however, lacks a general treatment and is challenging to establish even for specific languages.

>>>>

\paragraph{Dependent types}
Logical frameworks with bidirectional typing are proposed by \citet{Reed2008} and \citet{Felicissimo2023}.
\citeauthor{Felicissimo2023}'s framework is more expressive than \citeauthor{Reed2008}'s, due to its ability to specify rewriting rules.
Both frameworks extend \LF, enabling generic bidirectional type checking for dependent type theories.
They also incorporate notions of signatures and mode-correctness (called \emph{strictness} and \emph{validity}, respectively, in their contexts) but differ from ours in several ways.

First, the number of operations introduced by a signature in \LF is finite, so constructs like spine application seem impossible to define.
Second, \citeauthor{Reed2008} and \citeauthor{Felicissimo2023} deal with decorated raw terms only, while our theory bridges the gap between ordinary and decorated raw terms by mode decoration.
Lastly, \citeauthor{Felicissimo2023} classifies operations \emph{a priori} into introduction and elimination rules, and follows the Pfenning recipe assigning, for example, the synthesising mode to each elimination rule and its principal argument.
As pointed out by \citeauthor{Dunfield2021} that bidirectional typing is essentially about managing information flow, and that some systems remarkably deviate from this recipe, we do not enforce it but establish our results on any reasonable information flow.
